---
title: "MY470 Computer Programming: The R Language"
author: "Main contributors: Siân Brooke, Friedrich Geiecke"
date: "Week 8 Lab"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE}
knitr::opts_chunk$set(error = TRUE)
```

## R and Python

R and Python are two languages that you will often hear talked about in
Data Science. Both are open source (free to use and develop).

**Python**

-   General purpose language
-   Object-oriented (relies on classes)
-   Great for data structures and programming in general, it has a vast
    collection of libraries that you can use - but they can be tricky to
    install
-   Intuitive syntax and decent performance

**R**

-   Oriented to statistical analysis and data processing on small scale
-   Functional programming (uses functions)
-   It has a very huge collection of packages to do almost anything you
    might imagine with data (they are easy to install)
-   As a general programming language, it is much slower than Python and
    the syntax is arguably not as clear

## Python to R

There are many similarities between R and Python - but this can also be
a common source of error.

### Assignment and Arithmetic Operators

Python

``` python
# Assignment
a, b = 10, 20
# Power
a ** b
# Remainder (Modulus)
a % b
```

R

``` r
# Assignment
a <- 10; b <- 20
# Power
a ^ b
# Remainder (Modulus)
a %% b
```

### Logical Operators

An element-wise or simple logical operator will consider an entire
object whereas a short-circuit operator stops as soon as it evaluates a
test that produces a specified result.

For short circuit "and", no tests are evaluated after the first "false".

For short circuit "or", no tests are evaluated after the first "true".

Python

``` python
# Short-circuit AND
a and b
# Short-circuit OR
a or b
# Element-wise AND
a and b
# Element-wise OR
a or b
```

R

``` r
# Short-circuit AND
a && b
# Short-circuit OR
a || b
# Element-wise AND -> at the bit-level
a & b # do not use this in python (even if it works)
# Element-wise OR
a | b
```

### Sequences

Python

``` python
# 1, 2, [...] 10
range(1, 11) # if we want to start from 1 -> need to state it 
# List of numeric type
x = [2, 3, 0, 6]
# Updating at an index
x[i] = 5
```

R -\> here: everything is a vector (even if we just define a single
value)

``` r
# 1, 2, [...] 10
seq(10)
1:10
# List of numeric type
x <- c(1, 2.0, 4, 5)
# Updating at an index
x[i] <- 5
```

### Concatinate

`c()` is a generic function that combines its arguments.

By default it combines its arguments to form a vector. -\> even a single
element

``` r
c(1, 2, 3, 4, 5)
```

All arguments are coerced (forced) to a common type which is the type of
the returned value.

``` r
x <- c(1, 2, 3, "four")

> print(x)
# Output
"1", "2", "3", "four"

typeof(x)
# Output
"character"
```

All attributes (labeled values you can attach to an object) except names
(labels) are removed.

Note that numbers are stored as doubles (floats) by default in R. If you
want to save a number as an integer, you need to use the suffix `L`.
This is shown below.

``` r
x <- c(1L, 2L, 3L)
```

## Indexing

Let's say I'm working on a problem that involves a list of years:

```         
1993, 1994, 1995, 1996, 1997, 1998, 1999, 2000
```

I save a list of these dates in the variable `dates_list`.

-   **Q1.** In Python, what date would I get if I ran `dates_list[2]`
    -\> 1995 (index starts from 0)

-   **Q2.** What would I get in R for `dates_list[2]`? -\> 1994 (index
    starts from 1)

```{r}
# Lists in R are similar to dictionaries in Python

x <- c(1, 2, 3, 4, 5)

# we can but don't have to name them!
num_list <- list(one = x,
                 two = c("one", "two", "three"),
                 three = matrix(data = 1:6, nrow = 3, ncol = 3))
# NOTE: matrix() creates a matrix (table) from a given set of values

print(num_list)

# we get a warning:Warning: data length differs from size of matrix: [6 != 3 x 3]
# we try to put 6 numbers into a 3x3 matrix (has 9 spots)
# will get a warning but what r does it that it just cycles through the data
# means that it starts again at 1, 2, 3
```

## Indexing

```{r}
# We can index a list with brackets
# if it is a named list, it behaves like a dictionary in Python
print(num_list["two"])

# We can also index specific elements with $
# same as above, just different syntax
print(num_list$two)

# We can access data inside a list element by combining double and single brackets
# By using the double brackets, the list structure is dropped
print(num_list[["two"]])

print(class(num_list["two"]))
print(class(num_list[["two"]]))
```

It's easy to pull out specific entries in a vector using `[]`. For
example:

```{r}
# In Python, we would need to do all of this with looping!

student_names <- c("Bill", "Jane", "Sarah", "Fred", "Paul")
math_scores <- c(80, 75, 91, 67, 56)
verbal_scores <- c(72, 90, 99, 60, 68)

# Indexing
math_scores[3]

# Slicing (Inclusive)
math_scores[1:3]

# Remove by index
# "everything but" -> exclude (DIFFERENT FROM PYTHON!)
math_scores[-(4:5)]

# Filter by a condition
# subset math_scores where verbal_scores are larger than 90
# which leads to a true/false vector -> returns true or false
math_scores[which(verbal_scores >= 90)]

# Set a value
math_scores[3] <- 92

# Result
math_scores
```

## Data Frames

```{r}
# Idea: rows of data and columns of variables (different types)

# composing it as named object -> named columns and giving data as a vector
my_dataframe <- data.frame(title = c("Dr", "Prof", "Prof"),
                         lname = c("Strangelove", "Moriarty", "Farnsworth"),
                         favenum = c(13 , 99, 144))

# calling a column that does not exist = creating it
my_dataframe$fname <- c("", "James", "Hubert") 

print(my_dataframe)
```

## Control Flow

```{r}
# Similar to Python -> same logic but slightly different syntax
# R: indentation only for legibility, not functionality (like in Python)
# need to use () for if statement and {} to indicate what happens

x <- 5
y <- 10

# Indentation is not strictly necessary, but preferred for readability 

# The if code block is in rounded brackets
if (x < y) {
    print("x is smaller than y!")
    # the else if code block is also in rounded brackets
} else if (y < x) {
    print("y is smaller than x!")
} else {
    print("no number is smaller")
} # the code in the curly brackets will run if the conditional statement is true

```

```{r}
# For and while loops work pretty much the same way.

chr_vec <- c("this", "is", "how", "a", "for", "loop", "works")
x <- 5

# For loop
# works similar to python, just slightly different syntax
for (txt in chr_vec){
    print(txt)
}

# While loop
while (x < 10){
    print(x)
    x <- x + 1
}
```

```{r}
# Similar to range() in python ...

# length() is equivalent to range()

for (i in 1:length(chr_vec)) { # for i in range(0, len(chr_vec)) -> how it would be in Python
    print(i)
    print(chr_vec[i])
}
```

## Functions

```{r}
# Functions also make use of different brackets

# To read more about function documenantion in R, see:
# Need to use libraries so that comments are interpreted in a certain way
# in R: #' before a function will then be treated as a docstring (if used with e.g. roxygen2)


# https://cran.r-project.org/web/packages/roxygen2/vignettes/roxygen2.html
# https://cran.r-project.org/web/packages/docstring/vignettes/docstring_intro.html
# https://josephcrispell.github.io/2021/07/26/creating-R-docstring.html

#' Short title for function
#'
#' Longer description of the function

# we have the parameters in brackets
# can also define a default parameter (e.g. here: num_teachers: 1)
# also: example of how to use indentation to make it more legible:
class_fits <- function(num_teachers = 1, num_students, room_size) 
  {
    if (num_teachers + num_students < room_size)
      {
        print("Hooray! The room is big enough")
      } 
    else
      {
        print("Oh no, the room is too small") 
      }
  }

# We can supply a value for each argument
class_fits(2, 13, 18)

# Or not, because we supplied a default for teachers
class_fits(num_students = 18, room_size = 19)
```

### Pipe Operator

```{r, warning=FALSE}

# includes tools to keep everything more clean etc.
library(tidyverse)

# You might have noticed that your code in R often contains many parentheses. 
# When you have complex code, this will often mean that you will have to nest
# those parentheses together. This makes your R code hard to read and understand.
# The pipe operator is for this exact purpose.

x <- c(0.322, 0.237, 0.342, 0.983, 0.987 , 0.991, 0.129)

# Compute the logarithm of x, compute the exponential function, round the result
round(exp(log(x)), 1)

# With pipe this is:
# feeds into: feed x into log, feed result into exponent, feed that into rounding to 1
x %>% log() %>%
    exp() %>%
    round(1)

# NOTE: You don't need to include the brackets (i.e. log()) here,
# but doing so increases the legibility of your code.
```

## Recap: Terminology in R

**Vector** - Sequence of data elements of the same basic type. - A
vector is either logical, integer, double, complex, character, or raw
and can have any attributes except a dimension attribute - A
**vectorised operation** refers to the ability to do a mathematical
operation on a list -- or "vector" -- of numbers in a single step

**Lists** - Closest in python is a dictionary - Has a name, value
structure

**Matrix** - Arranges data from a vector into a table - Data has to be
the same type

**Data Frame** - A matrix-like R object in which the columns can be
different types (numeric, character, logical, etc.)

**Factors** - Used to represent categorical data - Can be ordered or
unordered and are an important class for statistical analysis and for
plotting

## Exercises

### Loop Exercises

Below are the for-loops from class in Week 3. For each loop, create a
chunk in R and translate the python code into R.

``` python
# Exercise 1: Create a list that contains all integers from 1 to 100 (inclusive), 
# except that it has None for every integer that is divisible by 3 
# Your list should look like: [1, 2, None, 4, 5, None, 7, 8, None, 10, ...]

ls = []
for i in range(1, 101):
    if i % 3 == 0:
        ls.append(None)
    else:
        ls.append(i)
print(ls)
```

```{r}
# equivalent of None in Python is NA
ls <- c(seq(100))

ls[which(seq(100)%%3 == 0)] <- NA

print(ls)


# solution from class -> with a for-loop (not very efficient!)
# rep replicates the values in x -> quick way to say that I want a vector with one element
# of type double
ls2 <- rep(0)
for (i in 1:100)
{
  if(i %% 3 == 0)
  {
    ls2[i] <- NA
  }
  else
  {
    ls2[i] <- i
  }
}
print(ls2)


# with using functions (more efficient!)
ls1 <- 1:100
# can tell it what the test is, what to do with yes and no
# we can call on the whole list because we are operating on vectors!
# basically the same as iterating in python
ls1 <- ifelse(ls1%%3==0, NA, ls1)
print(ls1)

```

*Your cell here.*

``` python
# Exercise 2: Sum the even integers from the list below.

lst = [1, 3, 2, 4.5, 7, 8, 10, 3, 5, 4, 7, 3.33]

summ = 0
for i in lst:
    if i % 2 == 0:
        summ += i
print(summ)
```

*Your cell here.*

```{r}
lst <- c(1, 3, 2, 4.5, 7, 8, 10, 3, 5, 4, 7, 3.33)

lst[which(lst%%2 != 0)] <- 0

summ <- sum(lst)

print(summ)


# solution from class:
# the expression gives true if the values satisfy the condition and then only sums those
sum(lst[lst%%2==0])


```

### Function Exercises

Translate the functions from the class exercises in Week 4 (below) into
R.

``` python
# Exercise 3

def zero_list(alist):
    """Takes a list and returns another list of the same length 
    that looks like [0, 0, 0, ...].
    """
    newlist = [0]*len(alist)  
    return newlist

mylist = [1, 2, 3]

zerolist = zero_list(mylist)  
print(mylist)
print(zerolist)
```

*Your cell here.*

```{r}
# Didn't get to it, so only solution from class:
# something didn't work?

zero_vect <- function(a)
{
  newvector <- integer(length(a))
  return newvector
}

v <- c(1, 2, 3)
zerov <- zero_vect(v)
print(zerov)
```

``` python
# Exercise 4

scientists = {'Alan Turing': 'mathematician', 'Richard Feynman': 'physicist',
              'Marie Curie': 'chemist', 'Charles Darwin': 'biologist',
              'Ada Lovelace': 'mathematician', 'Werner Heisenberg': 'physicist'}

def print_professions(dic):
    """Takes a dictionary of {Name: profession} and prints
    'Name was a profession.'
    """
    for i in dic:
        print(i + ' was a ' + dic[i] + '.')
        
print_professions(scientists)
```

```{r}
# from class:

scientists <- list('mathematician', 'physicist', 'chemist')
names(scientists) <- c('Alan Turing', 'Richard Feynman', 'Marie Curie')

# paste runs on all of the elements
paste(names(scientists), 'was a', scientists)

# result is a vector of three strings -> made each element one string


```

### Vectorizing Exercises

Vectorizing is a efficient alternative to for-loops in R. Vectorization
is the operation of converting repeated operations on simple numbers
(“scalars”) into single operations on vectors or matrices.

As we covered earlier, a vector is the elementary data structure in R
and is “a single entity consisting of a collection of things”.

Like **list comprehensions** in Python, many of the above loop
constructs can be made more efficient by using vectorization. At a lower
level, the alternative vectorized form translates into code which will
contain one or more loops in the lower-level language the former was
implemented and compiled (Fortran, C, or C++ ).

**We can vectorize in Python too**, particularly through the `map()`
function, which applies a given function to each item of a given
iterable.

Below we have the list comprehensions that we wrote in class in Week 3.
For each Python snippet, create a chunk in R and translate the code into
R.

``` python
# Exercise 5: Create a new list containing the squares of the integers 
# in the list below.

lst = [1, 3, 2, 4.5, 7, 8, 10, 3, 5, 4, 7, 3.33]
ans = [i**2 for i in lst if type(i) == int] 
```

```{r}
# as it is right now, the whole vector will be forced into 
# double because there is some doubles (4.5, 3.33)
lst <- c(1, 3, 2, 4.5, 7, 8, 10, 3, 5, 4, 7, 3.33)

# option 1:
lst[lst%%1==0]^2

# option 2: lst[!lst%%1]^2

```

``` python
# Exercise 6: Using the function ifelse(), replicate the list comprehension below

import numpy as np

x = np.random.normal(0, 1, 100)
y = ["Positive" if i > 0 else "Negative" for i in x]
```

```{r}
# didn't have time for this in class
```

```{r}
# Exercise 7: Vectorize the code below. This is a bad loop with 'growing' data.

# Set seed to compare the results
set.seed(42)

m <- 10
n <- 10

# Create matrix of normal random numbers
# replicate(num_of_replications, expression to evaluate)
# 10 x 10 normally distributed matrix
mymat <- replicate(m, rnorm(n))

# Transform into data frame
mydframe <- data.frame(mymat)

for (i in 1:m) {
  for (j in 1:n) {
    # For row, column, add some noise
    mydframe[i,j] <- mydframe[i,j] + 10*sin(0.75*pi)
  }
}
mydframe
```

## Apply Functions

The apply family functions in R are a well-known set of R functions that
allows you to perform complex tasks over arrays, avoiding the use of
for-loops. They run a loop in C (a lower level language), which means
they are pretty quick. You can do this in several ways, depending on the
value you specify to the MARGIN argument, that can be normally set to 1,
2 or c(1, 2).

can be very useful to apply to a whole dataset etc!

The syntax of the apply command is:

``` r
apply(X,       # Array, matrix or data frame
      MARGIN,  # 1: columns, 2: rows, c(1, 2): rows and columns
      FUN,     # Function to be applied
      ...)     # Additional arguments to FUN
```

```{r}
some_vector <- c(1, 2, 5)

# User defined function
my_function <- function(x) {
  x + 5
}

# `sapply` applies a function to every element of a vector
sapply(some_vector, my_function)

# Anonymous function - does not require you to define a named function
sapply(some_vector, function(y) y + 5)
```

### Apply Exercises

Using the sample data specified below, use `?` with `apply()` to
complete the steps below:

-   Exercise 8: Create a column which is the mean of each row.
-   Exercise 9: Create a row which is the sum of each column.
-   Exercise 10: Create a function named `fun` that calculates the
    square of a number and convert the output to character if the
    character argument is set to `TRUE`.

```{r}

# Data frame
sample.df <- data.frame(x = 1:4, y = 5:8, z = 10:13)

```
